DOpElib: Differential Equations and Optimization Environment; A Goal Oriented Software Library for Solving PDEs and Optimization Problems

### 2012-11-09 Paper_May_2012/main.tex
Die abgesprochenen Zitierungsaenderungen und jetzt auch ein Zitat von der DOpElib Webseite

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exist a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \cite{dealnew}, dune \cite{dune}, 
FEniCS \cite{fenics}
or commercial solvers like ANSYS (with its Fluent package) \cite{ansys},
COMSOL \cite{comsol} or simufact forming \cite{simufact}.
Finally, often, a lot of work is (still) done in Matlab \cite{matlab}.
In fact, these libraries provide very useful tools 
for a huge variety of applications.

@@

The presented toolkit \dope{}: 
\textit{Differential Equations and Optimization Environment}
fills this gap. We present a software
to solve both PDE problems as well as optimization problems 
constrained by a PDE. 
Its main feature is to give a unified interface to high level algorithms such as 
time-stepping methods, nonlinear solvers and optimization routines. 
\dope{} is designed such that the user only needs to write those parts
of the code that are problem dependent while all invariant 
parts of the algorithms
are reusable without any need for further coding.
In particular, the user is enabled to switch between various different 
algorithms without the need to rewrite the problem dependent code, 
though obviously he or she will
have to replace the algorithm object with an other one. 
This replacement can be done by replacing the appropriate object at only
one point in the code.

@@

The project was initially initiated at the University of Heidelberg in 2010 and  
at the present stage the following key features are supported by the library
\begin{itemize}
\item Solution of nonlinear stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various, easily exchangeable, 
  time-stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item Use of all finite elements of from \deal{} including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs by elimination of 
   the state variable. \cite{NoWr00} 
   \marginpar{spaeter zitieren}
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other control and state constraints.
\item Several examples showing how to solve various kinds of optimization 
  problems involving PDE constraints.
\item Goal-oriented mesh adaptation with the Dual Weighted Residual method, 
  following \cite{BR03}.
\item Different spatial triangulations for control and state variables without
  the need to reimplement the user dependent code when switching from one common 
  triangulation to different ones.
\end{itemize}

@@

\subsection{Solving of optimization problems}\label{sec:opt}
Upgrading the framework for solving optimization problems
is described in this section. A prototypical problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.
To keep the presentation brief we will neglect the additional bounds
\[
a \le q \le b,\qquad g(q,u) \le 0
\]
although they can be treated by \dope{} as well.
Then, in order to minimize the functional $J$ one may apply the, so-called, 
reduced approach, in which we eliminate the dependent variable $u$ by the solution 
operator $S$ of the PDE, i.e., 
\begin{align*}
a(q,S(q))(\phi) = 0 \quad \forall \phi\in V. 
\end{align*}
This leads to the problem
\begin{align*}
\min\;&j(q) = J(q,S(q)). 
\end{align*}
Then, one can apply any algorithm to solve the, unconstraint, minimization 
problem. This, will in any case require the evaluation of $j(q)$ which 
we can do, e.g., using \texttt{StatPDEProblem::ComputeReducedFunctional} for 
a stationary PDE. If one would like to use derivative based methods, one 
needs in addition derivatives of $j$ and thus of $J$ as well as of $S$. 
To do so, we remark, that one can calculate the derivative of 
$j$ using the adjoint calculus. Then the directional derivative 
of $j$ in direction $\delta q$ is given by
\[
j'(q)(\delta q) = J'_q(q,S(q))(\delta q) + (S^* J_u'(q,S(q)),\delta q)
\]
where the index $q$ or $u$ denote directional derivatives for the 
respective argument. To calculate action of the adjoint operator $S^*$ 
we note that $z =  S^* J_u'(q,S(q))$
can be computed via the adjoint PDE
\begin{equation}
  \label{opt_KKT_system}
  \begin{aligned}
    a_{u}'(q,u)(\phi , z) &= J'_{u}(q,u)(\phi) & \forall \phi&\in {\cal X}.
  \end{aligned}
\end{equation}
Similarly formulas for the second derivatives can be obtained.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{introduction}
Today, there exist a broad variety of software libraries
that provide numerical tools for solving partial differential
equations (PDEs). Prominent examples are 
\deal{} \cite{dealnew}, dune \cite{dune}, 
FEniCS \cite{fenics}
or commercial solvers as for instance ANSYS (with its Fluent package),
COMSOL, or simufact forming.
Finally, often, a lot of work is (still) done in Matlab.
In fact, these libraries provide very useful tools 
for a huge variety of applications.

@@

The presented toolkit \dope{}: 
\textit{Differential Equations and Optimization Environment} \cite{dope}
fills this gap. We present a software
to solve both PDE problems as well as optimization problems 
constrained by a PDE. 
Its main feature is to give a unified interface to high level algorithms such as 
time-stepping methods, nonlinear solvers and optimization routines. 
\dope{} is designed such that the user only needs to write those parts
of the code that are problem dependent while all invariant 
parts of the algorithms
are reusable without any need for further coding.
In particular, the user is enabled to switch between various different 
algorithms without the need to rewrite the problem dependent code, 
though obviously he or she will
have to replace the algorithm object with an other one. 
This replacement can be done by replacing the appropriate object at only
one point in the code.

@@

The project was initially initiated at the University of Heidelberg in 2010 and  
at the present stage the following key features are supported by the library
\begin{itemize}
\item Solution of nonlinear stationary and nonstationary PDEs in 1d, 2d, and 3d.
\item Various, easily exchangeable, 
  time-stepping schemes (based on finite differences), 
  such as forward Euler, backward Euler,
  Crank-Nicolson, shifted Crank-Nicolson, and Fractional-Step-$\Theta$ scheme.
\item Use of all finite elements of from \deal{} including hp-support.
\item Several examples showing the solution of several PDEs including
   Poisson, Navier-Stokes, Plasticity and fluid-structure interaction problems. 
\item Line search and trust region newton algorithms for the 
   solution of optimization problems with PDEs by elimination of 
   the state variable.
\item Interface to SNOPT for the solution of optimization problems with PDEs and
  additional other control and state constraints.
\item Several examples showing how to solve various kinds of optimization 
  problems involving PDE constraints.
\item Goal-oriented mesh adaptation with the Dual Weighted Residual method and
  standard residual based error estimators.
\item Different spatial triangulations for control and state variables without
  the need to reimplement the user dependent code when switching from one common 
  triangulation to different ones.
\end{itemize}

@@

\subsection{Solving of optimization problems}\label{sec:opt}
Upgrading the framework for solving optimization problems
is described in this section. A prototypical problem reads:
\begin{align*}
\min\;&J(q,u) \\
  &\text{s.t.}\; a(q,u)(\phi) = 0 \quad \forall \phi\in V,\\
  &a \le q \le b,\\
  &g(q,u) \le 0,  
\end{align*}
where $u$ is a FE-function and $q$ can either be a FE-function or some 
fixed number of parameters, $a$ and $b$ are constraint bounds for the control $q$,
and $g(\cdot)$ is some state constraint.
To keep the presentation brief we will neglect the additional bounds
\[
a \le q \le b,\qquad g(q,u) \le 0
\]
although they can be treated by \dope{} as well.
Then, in order to minimize the functional $J$ one may apply the, so-called, 
reduced approach, in which we eliminate the dependent variable $u$ by the solution 
operator $S$ of the PDE, i.e., 
\begin{align*}
a(q,S(q))(\phi) = 0 \quad \forall \phi\in V. 
\end{align*}
This leads to the problem
\begin{align*}
\min\;&j(q) = J(q,S(q)). 
\end{align*}
Then, one can apply any algorithm to solve the, unconstraint, minimization 
problem. This, will in any case require the evaluation of $j(q)$ which 
we can do, e.g., using \texttt{StatPDEProblem::ComputeReducedFunctional} for 
a stationary PDE. If one would like to use derivative based methods, one 
needs in addition derivatives of $j$ and thus of $J$ as well as of $S$. 
To do so, we remark, that one can calculate the derivative of 
$j$ using the adjoint calculus. Then the directional derivative 
of $j$ in direction $\delta q$ is given by
\[
j'(q)(\delta q) = J'_q(q,S(q))(\delta q) + (S^* J_u'(q,S(q)),\delta q)
\]
where the index $q$ or $u$ denote directional derivatives for the 
respective argument. To calculate action of the adjoint operator $S^*$ 
we note that $z =  S^* J_u'(q,S(q))$
can be computed via the adjoint PDE
\begin{equation}
  \label{opt_KKT_system}
  \begin{aligned}
    a_{u}'(q,u)(\phi , z) &= J'_{u}(q,u)(\phi) & \forall \phi&\in {\cal X}.
  \end{aligned}
\end{equation}
Similarly formulas for the second derivatives can be obtained, see, e.g., 
\cite{BeMeVe06}.

@@

Based on this classes we can now solve the optimization problem, for instance using 
a Newton type method for the reduced problem, see, e.g., \cite{NoWr00}. This yields 
an object \texttt{ReducedNewtonAlgorithm} which needs to be initialized 
with the \texttt{OptProblemContainer} and then provides 
the following method to be called by the user
\begin{lstlisting}
  ReducedNewtonAlgorithm::Solve(ControlVector<VECTOR>& )
\end{lstlisting}
where the argument provides an initial guess for the optimal control.
